Methodology for Assessing the State of the Practice for Domain X

### 2021-10-20 StateOfPractice/Methodology/Methodology.tex
Updates to Methodology document

\usepackage[round]{natbib}
\usepackage{booktabs}
\usepackage{amsmath,amsthm}
\usepackage{hyperref}
\usepackage{xcolor}

@@

\theoremstyle{definition}
\newtheorem{defn}{Definition}

@@

\begin{abstract}
	...
\end{abstract}

@@

Given the importance of research software, scientists and engineers are pushing
for methods and tools to sustainably develop hiqh quality software.  This is
evident from the existence of such groups as the Software Sustainability
Institute (\href{https://www.software.ac.uk/} {SEI}) and Better Scientific
Software (\href{https://bssw.io/} {BSS}).  Sustainability promoting groups such
as these are necessary because unfortunately the current state-of-the-practice
for research software often does not incorporate state-of-the-art Software
Engineering (SE) tools and methods \citep{JohansonAndHasselbring2018}.  The lack
of SE tools and methods contributes to sustainability and reliability problems
\citep{FaulkEtAl2009}.  Problems with the current state of the practice are
evident from embarrassing failures, like a retraction of derived molecular
protein structures \citep{Miller2006}, false reproduction of sonoluminescent
fusion \citep{PostAndVotta2005}, and fixing and then reintroducing the same
error in a large code base three times in 20 year
\citep{MilewiczAndRaybourn2018}.  To improve this situation, we need to first
fully understand the current state of the practice for research software.

The purpose of our proposed methodology is to understand how software quality is
impacted by the software development principles, processes and tools currently
used within research software communities.  Since research software is so broad
a category, we will reduce the scope of our methodology to focus on one specific
research software domain at a time.  This ``state of the practice for domain X''
exercise will build off of prior work on measuring/assessing the state of
software development practice in several research domains.  We will update the
work that was done previously for domains such as Geographic Information Systems
\citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
Seismology software \citep{SmithEtAl2018}, and Statistical software for
psychology \citep{SmithEtAl2018_StatSoft}.

In the previous ``state of the practice'' project, we measured 30 software
projects for each domain, but the measures were relatively shallow.  With this
re-boot we will still target about 30 software examples from each domain, but we
will be collecting more data by adding more measures and experiments.  In
keeping with the previous project, we still have the constraint that the work
load for applying the methodology to a given domain needs to be feasible for a
team as small as one individual, and for a time that is short, ideally 3 to 4
months of part-time work per domain.

To start with, the 
\href{https://github.com/adamlazz/DomainX/blob/master/TemplateToGradeSCSoft.pdf}
{previous set of questions} were critically assessed and modified.  In addition,
the following data is part of the new methodology:

\begin{itemize}
\item Characterization of the functionality provided by the software in the
  domain via a commonality analysis.
\item Usability testing for a sample of software in the domain.
\item Empirical software engineering related data, such as the number of
  files, number of lines of code, percentage of issues that are closed, etc. 
\end{itemize}

Unlike the previous measurement process, the new methodology will involves and
engages a domain expert partner throughout.  We did not previously engage the
domain expert with the rationale that we wished to exclude potential bias.
However, this advantage is not worth not being able to evaluate the
functionality/usability of the software.  Moreover, not having an expert makes
publication more difficult, since there is no one to provide advice on how best
to approach journals and publishers.

In the proposed methodology, the collected data will be combined to rank the
software within each domain using the Analytic Hierarchy Process (AHP)
\citep{Saaty1980}.  As in the past process, we will use AHP to develop a list of software
ranked by quality.  However, in the new process we will not stop with this
list.  The domain expert will be consulted to verify the ordering, and to
discuss the decisions that led to the ranking.  The AHP process will be used to
facilitate a conversation with the domain experts as a means to deepen our
understanding of the software in the domain, and the needs of typical developers.

GitHub will be used to coordinate the work of the team of people that will
be involved in this project.  In addition to the project record left on GitHub,
the final data will be exported to Mendeley (as was done for the previous
quality measurement exercise, for instance the grades for GIS software are
summarized at \href{https://data.mendeley.com/datasets/6kprpvv7r7/1}
{https://data.mendeley.com/datasets/6kprpvv7r7/1}.)

The scope of this methodology includes observations on product, artifact and
process quality for research softwares. We leave the assessment of research
software using performance benchmarks to other projects, such as the work of
\cite{kaagstrom1998gemm}.

Note the following formatting conventions of this document. \textcolor{red}{Red}
text denotes a link to sections of this document. \textcolor{cyan}{Cyan}
text denotes a URL link. \textcolor{blue}{Blue} text denotes a citation.

\section{Research Questions} \label{ResearchQuestions}

@@

%In general questions:
%
%\begin{enumerate}
%\item Comparison between domains
%\item How to measure qualities
%\item How does the quality compare for projects with the most resources to %those
%  with the fewest?
%\item What skills/knowledge are needed by future developers?
%\item How can the development process be improved?
%\item What are the common pain points?
%\end{enumerate}
%
%For each domain questions:
%
%\begin{enumerate}
%\item Best examples within the domain
%\item What software artifacts?
%\item What are the pain points?
%\item Any advice on what can be done about the pain points?
%\end{enumerate}
%
%Measure the effort invested and the reward.  Related to sustainability.
%
%Collect the data and see what conclusions follow.  For an individual domain,
%between domains.  The process isn't so much about ranking the software as it is
%about looking at the software closely and see what conclusions arise.  The
%measurements are intended to force scrutiny, from different perspectives.

@@

To answer the above research questions (Section~\ref{ResearchQuestions}), we
systematically measure the quality of the software through data collection and a
series of experiments.  An overview of the measurement process is given in the
following steps, starting from determining a domain that is suitable for
measurement:

\begin{enumerate}
\item Identify the domain. (Section~\ref{SecIdentifyDomain})
\item \emph{Domain Experts}: Create a top ten list of software packages in the
  domain. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
\item Brief the Domain Experts on the overall objective, research proposal,
  research questions, measurement template, survey for short list projects,
  usability tests, maintainability
  experiments. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
\item Identify broad list of candidate software packages in the
  domain. (Section~\ref{SecIdentifyCandSoft})
\item Preliminary filter of software packages
  list. (Section~\ref{SecInitialFilter})
\item \emph{Domain Experts}: Review domain software
  list. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
\item Domain Analysis. (Section~\ref{SecDomainAnalysis})
\item \emph{Domain Experts}: Vet domain
  analysis. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
\item Gather source code and documentation for each prospective software
  package.
\item Collect empirical measures. (Section~\ref{SecEmpiricalMeasures})
\item Measure using measurement template. (Section~\ref{SecShallowMeasure})
\item Survey developers
(\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Questions
	to Developers.pdf}{Questions to Developers})
\item Use AHP process to rank the software packages. (Section~\ref{SecAHP})
\item Identify a short list of top software packages, typically four to six, for
  deeper exploration according to the AHP rankings of the measurements.
\item \emph{Domain Experts}: Vet AHP ranking and short
  list. (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Meeting
    Agenda withmDomainmExperts.pdf}{Meeting Agenda with Domain Experts})
\item With short list:
\begin{enumerate}
\item Usability experiments
  (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
\item Modifiability experiments
  (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})
\end{enumerate}
\item Rank short list. (Section~\ref{SecRankShortList}) \pmi{To be completed -
    add pairwise comparison tool}
\item Document answers for research questions.
\end{enumerate}

@@

\section{How to Identify the Domain} \label{SecIdentifyDomain}

A domain of research software must be identified. Research software is defined
in this exercise as ``software that is used to generate, process or analyze
results that [are intended] to appear in a publication'' \citep{hettrick2014uk}
in a scientific or engineering context.  Research software is a more general
term that covers what is often called scientific computing.  To be applicable
for the methodology described in this document, the chosen domain must have the
following properties:

@@

\section{How to Identify Candidate Software} \label{SecIdentifyCandSoft}

Once the domain of interest is identified, the candidate software for measuring
can be found through search engine queries targeting authoritative lists of
software.  Potential places to search include \href{https://github.com/}
{GitHub}, \href{https://swmath.org/} {swMATH} and domain related publications,
such as review articles. Domain experts are also asked for their suggestions and
are asked to review the initial draft of the software list.

@@

\begin{enumerate}
\item The software functionality must fall within the identified domain.
\item The source code must be viewable.
\item The empirical measures listed in Section \ref{SecEmpiricalMeasures} should
  ideally be available, which implies a preference for GitHub-style repositories.
\item The software cannot be marked as incomplete or in an initial development
  phase.
\end{enumerate}

\section{How to Initially Filter the Software List} \label{SecInitialFilter}

@@

Copies of both the initial and filtered lists should be kept for traceability purposes.

@@

The final result of the domain analysis will be
tables of commonalities, variabilities, and parameters of variation of a program
family.   \cite{smith2008commonality} present a template for conducting a
commonality analysis, which was referred to when conducting this
work. \cite{weiss1998commonality} describes another commonality analysis
technique for deciding the members of a program family which readers are
encouraged to be familiar with. \cite{SmithAndChen2004} and
\cite{SmithMcCutchanAndCarette2017} are examples of a commonality analysis for a
family of mesh generating software and a family of material models,
respectively. The steps to produce a commonality analysis are:

\begin{enumerate}
\item Write an Introduction
\item Write the Overview of Domain
\item List Commonalities
\item List Variabilities
\item List Parameters of Variation
\item Add Terminology, Definitions, Acronyms
\end{enumerate}

A sample commonality analysis for Lattice Boltzmann Solvers can be found
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Peter-Notes/Commonality-Analysis-LB-Systems.pdf}
{here}.

\section{Empirical Measures} \label{SecEmpiricalMeasures}

Some quality measurements rely on the gathering of raw and processed empirical
data. We focus on data that is reasonably easy to collect, which we combine and
analyze for this assessment. The measures that are collected relate to the
research questions (Section~\ref{ResearchQuestions}). For instance, we collect some of the
data to see how large a project is. Other measures intend to ascertain a
project’s popularity, or how active the project is.

Section \ref{rawdata} orients the reader on what raw data is collected.  Some of this data can be observed from GitHub repository metrics. The rest can be collected using freeware tools. \href{https://github.com/tomgi/git_stats}{GitStats} is used to measure
the number of binary files as well as the number of added and deleted lines in a
repository. This tool is also used to measure the number of commits over
different intervals of time. \href{https://github.com/boyter/scc}{Sloc Cloc and
  Code (scc)} is used to measure the number of text based files as well as the
number of total, code, comment, and blank lines in a repository. These tools
were selected due to their installability, usability, and ability to gather the
empirical measures listed below. Details on installing and running the tools can
be found in our 
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/A
  Guide to Empirical Measures.pdf} {Guide to Empirical Measures}. Section \ref{processeddata}
introduces the required processed data, which is calculated using the raw data.

@@

\begin{itemize}
\item Status of software package (dead or alive). Alive is defined as the
  presence of repository commits or software package version releases in the
  last 18 months.
\item Percentage of identified issues that are closed.
\item Percentage of code that is comments.
\end{itemize}

\section{Measure Using Measurement Template} \label{SecShallowMeasure}

@@

The
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}
{Measurement Template} is used to track measurements and quality scores for all
of the software packages in the domain. For each software package, we fill out one
column of the template. This process can take between 2 to 5 hours for each
package.  Project developers can be contacted for help regarding installation,
if necessary, but a cap of about 2 hours should be imposed on the entire
installation process, to keep the overall measurement time feasible.

@@

\begin{enumerate}
\item Gather the summary information into the top section of the document
  (Figure \ref{measurement_template_image})
\item Using the GitStats tool that is described in Section
  \ref{SecEmpiricalMeasures} gather the measurements for the Repo Metrics -
  GitStats section found near the bottom of the document
\item Using the SCC tool that is also described in Section
  \ref{SecEmpiricalMeasures} gather the measurements for the Repo Metrics - SCC
  section found near the bottom of the document
\item If the software package is found on git, gather the measurements for the
  Repo Metrics - GitHub section found near the bottom of the document
\item Review installation documentation and attempt to install the software
  package on a virtual machine
\item Gather the measurements for installability
\item Gather the measurements for correctness and verifiability
\item Gather the measurements for surface reliability
\item Gather the measurements for surface robustness
\item Gather the measurements for surface usability
\item Gather the measurements for maintainability
\item Gather the measurements for reusability
\item Gather the measurements for surface understandability
\item Gather the measurements for visibility and transparency
\item Assign a score out of ten for each quality. The score can be measured
  using the
  \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/MeasurementTemplate_ImpressionCalculator.xlsx}
  {Measurement Template Impression Calculator}. For each quality measurement,
  the file indicates the appropriate score to assign the measurement based on
  possible measurement values.
\end{enumerate}

@@

The Analytical Hierarchy Process (AHP) is a decision-making technique that can
be used when comparing multiple options by multiple criteria. In our work AHP is
used for comparing and ranking the software packages of a domain using the
quality scores that are gathered in the
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Combined_MeasurementTemplate_EmpiricalMeasures.xlsx}
{Measurement Template}. AHP performs a pairwise analysis between each of the
quality options using a matrix which is then used to generate an overall score
for each software package for the given criteria. \cite{SmithEtAl2016} shows how
AHP is applied to ranking software based on quality measures. We have developed
a tool for conducting this process. The tool includes an AHP JAR script and a
sensitivity analysis JAR script that is used to ensure that the software package
rankings are appropriate with respect to the uncertainty of the quality
scores. The README file of the tool is found
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{here}. This
file outlines the requirements for, and configuration and usage of, the JAR
scripts. The JAR scripts, source code, and required libraries are located in the
same folder as the README file.

\section{Rank Short List} \label{SecRankShortList}

An AHP pairwise comparison between the short list packages with respect to
usability and modifiability experiment results is conducted, and then reviewed by the domain expert. The tool can be found here. The usability and modifiability
survey can be found
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}
{here}. This step is intended to start a conversation about the relative quality
of short-listed software packages.

@@

>>>>

\usepackage[round]{natbib}
\usepackage{booktabs}
\usepackage{amsmath,amsthm}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{longtable}
\usepackage{paralist}
\usepackage{totcount}
\usepackage{booktabs}

@@

\newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\small\bf
+}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\small\bf +}}

\newcounter{totHours} %to track the estimate of total hours to assess SOP
\setcounter{totHours}{0}
\regtotcounter{totHours}

@@

\begin{abstract}
	To improve software development methods and tools for research software, we
	first need to understand the current state of the practice.  Therefore, we
	have developed a methodology for assessing the state of the software
	development practices for a given research software domain.  The methodology
	is applied to one domain at a time in recognition that software development in
	different domains is likely to have adopted different best practices.
	Moreover, providing a means to measure different domains facilitates
	comparison of development practices between domains.  For each domain we wish
	to answer questions such as: 
  \begin{inparaenum}[i)]
    \item What artifacts (documents, code, test cases, etc.) are present?
    \item What tools are used?
    \item What principles, process and methodologies are used?
    \item What are the pain points for developers?
    \item What actions are used to improve qualities like maintainability and
    reproducibility?
  \end{inparaenum} 
  To answer these questions, our methodology prescribes the following steps: 
  \begin{inparaenum}[i)] 
    \item Identify the domain;
    \item Identify a list of candidate software packages;
    \item Filter the list to a length of about 30 packages;
    \item Gather source code and documentation for each package;
    \item Collect raw data on each software repo, like number of stars, number
    of open issues, number of lines of code, and so on;
    \item Fill in the measurement template (the template consists of 108 questions
    to assess 9 qualities (like installability, usability and visibility));
    \item Interview developers (the interview consists of 20 questions and takes
    about an hour);
    \item Rank the software using the Analytic Hierarchy Process (AHP); and,
    \item Analyze the data to answer the questions posed above.
  \end{inparaenum}
  A domain expert should be engaged throughout the process, to ensure that
  implicit information about the domain is properly represented and to assist
  with conducing an analysis of the commonalities and variabilities between the
  30 selected packages.  Using our methodology, spreadsheet templates and AHP
  tool, we estimate (based on our experience with using the process) the time to
  complete an assessment for a given domain at \total{totHours} person hours.
\end{abstract}

@@

Given the importance of research software, scientists and engineers are pushing
for methods and tools to sustainably develop high quality software.  This is
evident from the existence of such groups as the Software Sustainability
Institute (\href{https://www.software.ac.uk/} {SEI}) and Better Scientific
Software (\href{https://bssw.io/} {BSS}).  Sustainability promoting groups such
as these are necessary because unfortunately the current state-of-the-practice
for research software often does not incorporate state-of-the-art Software
Engineering (SE) tools and methods \citep{JohansonAndHasselbring2018}.  The lack
of SE tools and methods contributes to sustainability and reliability problems
\citep{FaulkEtAl2009}.  Problems with the current state of the practice are
evident from embarrassing failures, like a retraction of derived molecular
protein structures \citep{Miller2006}, false reproduction of sonoluminescent
fusion \citep{PostAndVotta2005}, and fixing and then reintroducing the same
error in a large code base three times in 20 year
\citep{MilewiczAndRaybourn2018}.  To improve this situation, we need to first
fully understand the current state of the practice for research software
development.

The purpose of our proposed methodology is to understand how software quality is
impacted by the software development principles, processes and tools currently
used within research software communities.  Since research software is so broad
a category, we will reduce the scope of our methodology to focus on one specific
research software domain at a time.  To emphasize that the method is generic, we
will label the specific domain as X within this document.

This ``state of the practice for domain X'' exercise builds off of prior work on
measuring/assessing the state of software development practice in several
research domains.  We have updated the work that was done previously for domains
such as Geographic Information Systems \citep{SmithEtAl2018_arXivGIS}, Mesh
Generators \citep{SmithEtAl2016}, Seismology software \citep{SmithEtAl2018}, and
Statistical software for psychology \citep{SmithEtAl2018_StatSoft}.  Initial
tests of the new methodology have been done for medical image analysis software
\citep{Dong2021} and for Lattice Boltzmann Method (LBM) software
\citep{Michalski2021}.

In the previous ``state of the practice'' project, we measured 30 software
projects for each domain, but the measures were relatively shallow.  With this
re-boot we still target about 30 software examples from each domain, but we are
now collecting more data.  In keeping with the previous project, we still have
the constraint that the work load for applying the methodology to a given domain
needs to be feasible for a team as small as one individual, and for a time that
is short, ideally under a person month per domain.\footnote{A person month is
considered to be $20$ working days ($4$ weeks in a month, with $5$ days of work
per week) at $8$ person hours per day, or $20 \cdot 8 = 160$ person hours.}

To begin our re-boot of the previous methodology we critically assessed, and
subsequently modified, our
\href{https://github.com/adamlazz/DomainX/blob/master/TemplateToGradeSCSoft.pdf}
{previous set of measures}.  In addition, the following data has been added to
the new methodology:

\begin{itemize}
\item Characterization of the functionality provided by the software in the
  domain via a commonality analysis.
\item Project repository related data, such as the number of files, number of
  lines of code, percentage of issues that are closed, etc. 
\item Interviews with software developers in domain X.
%\item Usability testing for a sample of software in the domain.  
\end{itemize}

Unlike the previous measurement process, the new methodology involves and
engages a domain expert partner throughout.  We did not previously engage the
domain expert with the rationale that we wished to eliminate potential bias.
However, this advantage is not worth the inability to evaluate the functionality
of the software.  Moreover, not having an expert makes navigating the on-line
resources difficult, since some information will be implicit to domain experts,
and not every statement found on-line will be accurate   The importance of the
domain expert is particularly noteworthy when it comes time to publication and
dissemination of the state of the practice assessment.  Throughout this document
the person (or persons) that provide domaine expertise will be designated as the
\textit{Domain Expert}.

In the proposed methodology, the collected data is combined to rank the software
within each domain using the Analytic Hierarchy Process (AHP) \citep{Saaty1980}.
As in the previous measurement exercise, we use AHP to develop a list of
software ranked by quality.  However, in the new process we do not stop with
this list.  The Domain Expert is consulted to verify the ordering, and to
discuss the decisions that led to the ranking.  The AHP process is used to
facilitate a conversation with the Domain Expert as a means to deepen our
understanding of the software in the domain, and the needs of typical
developers.

So that the collected data for a given domain can benefit the scientific
community, our recommendation is that all collected data be made public.  For
instance, the data collection for each domain could be put on a GitHub
repository.  In addition to the project record left on GitHub, the final data
can be exported to \href{https://data.mendeley.com/}{Mendeley Data}.  As an
example, the measurements for the
\href{https://data.mendeley.com/datasets/6kprpvv7r7/1} {state of practice on GIS
software} are available on Mendeley.  Ideally the full analysis of the state of
the practice for domain X will be published in a suitable journal.

The scope of this methodology includes observations on product, artifact and
process quality for research softwares. We leave the assessment of research
software using performance benchmarks to other projects, such as the work of
\cite{kaagstrom1998gemm}.  Currently we are also leaving experiments to measure
usability and modifiability as future work, as discussed in
Section~\ref{SecFutureWork}.

Within this document the following typographic conventions have been adopted:
\begin{inparaenum}[i)] 
  \item \textcolor{red}{Red} text denotes a link to sections of this document. 
  \item \textcolor{cyan}{Cyan} text denotes a URL link.
  \item \textcolor{blue}{Blue} text denotes a citation.
\end{inparaenum}

The full methodology is presented in the sections that follow.
Section~\ref{SecResearchQuestions} highlights the research questions that are to
be answered for each measured domain. These questions are answered by the data
collected using the process outlined in Section~\ref{StepsAQDS}.  The major
steps in the process are outlined in
Sections~\ref{SecIdentifyDomain}--\ref{SecAHP}.  Following this, the time
required for assessing a single domain is given in
Section~\ref{SecEstTimeRequired}.

\section{Research Questions} \label{SecResearchQuestions}

@@

@@

To answer the above research questions (Section~\ref{SecResearchQuestions}), we
systematically measure the quality of the software through data collection and a
series of experiments.  An overview of the measurement process is given in the
following steps, starting from determining a domain that is suitable for
measurement:

\begin{enumerate}
\item Identify the domain (X). (Section~\ref{SecIdentifyDomain})
\item Ask the Domain Expert to create a top ten list of software packages in the
  domain. (Section~\ref{SecDomainExpert})
\item Meet with the Domain Expert to brief them on the overall objective,
  research proposal, research questions, measurement template, and developer
  interviews. (Section~\ref{SecDomainExpert}) \label{StepMeeting}
\item Identify broad list of candidate software packages in the domain.
  (Section~\ref{SecIdentifyCandSoft})
\item Preliminary filter of software packages list.
  (Section~\ref{SecInitialFilter})
\item Review software list with Domain Expert. (Section~\ref{SecDomainExpert})
\item Domain Analysis. (Section~\ref{SecDomainAnalysis})
\item Ask Domain Expert to vet domain analysis.
  (Section~\ref{SecDomainExpert})
\item Gather source code and documentation for each prospective software
  package.
\item Collect repository based information. (Section~\ref{SecEmpiricalMeasures})
\item Measure using measurement template. (Section~\ref{SecMeasureTemplate})
\item Survey developers (Section~\ref{SecSurveyQuestions})
\item Use AHP process to rank the software packages. (Section~\ref{SecAHP})
% \item Identify a short list of top software packages, typically four to six, for
%   deeper exploration according to the AHP rankings of the measurements.
\item Ask Domain Expert to vet AHP ranking. (Section~\ref{SecDomainExpert})
%\item Rank short list. (Section~\ref{SecRankShortList})
\item Answer research questions (from Section~\ref{SecResearchQuestions}) and
document answers.
\end{enumerate}

@@

\section{How to Identify the Domain?} \label{SecIdentifyDomain}

A domain of research software must be identified. Research software is defined
in this exercise as ``software that is used to generate, process or analyze
results that [are intended] to appear in a publication'' \citep{hettrick2014uk}
in a scientific or engineering context.  Research software is a more general
term for what is often called scientific computing.  To be applicable for the
methodology described in this document, the chosen domain must have the
following properties:

@@

\section{Interaction With Domain Expert} \label{SecDomainExpert}

As mentioned in the introduction (Section~\ref{SecIntroduction}), the Domain
Expert is an important member of the state of the practice assessment team.
Pitfalls exist if non-domain experts attempt to acquire an authoritative list of
software, or perform a commonality analysis or try to rank the software. The
main source of problems for non-domain experts is that they can only rely on
information that is available on-line, but on-line data has two potential
problems:
\begin{inparaenum}[i)]
  \item the on-line resources could have false or inaccurate information; and,
  \item the on-line resources could leave out relevant information that is so
in-grained with experts that nobody thinks to explicitly record the information.
\end{inparaenum}

Domain experts may be recruited from academia or industry.  The only
requirements are knowledge of the domain and a willingness to be engaged in the
assessment process.  The domain expert does not have to be a software developer,
but they should be a user of domain software.  Given that the domain experts are
likely to be busy people, the measurement process needs to benefit from their
expertise, without putting too much of a burden on their time.

In advance of the first meeting with the Domain Expert (Step~\ref{StepsAQDS} in
Section~\ref{StepsAQDS}) the expert is asked to create a top ten list of
software packages in the domain.  This is done to help the expert get in the
right mind set in advance of the first meeting.  Moreover, by doing the exercise
in advance, we avoid the potential pitfall of the expert approving the
discovered list of software without giving it adequate thought.  The emphasis
during the first meeting is for the Domain Expert to learn what is expected of
them.  The discussion should also cover avenues for publication and
dissemination.

The Domain Experts provide are asked to vet the collected data and analysis.  In
particular, they are asked to vet the proposed list of software packages, the
domain analysis and the AHP ranking.  These interactions can be done either
electronically or with in-person (or virtual) meetings.

\section{How to Identify Candidate Software?} \label{SecIdentifyCandSoft}

Once the domain of interest is identified, the candidate software for measuring
can be found through search engine queries targeting authoritative lists of
software.  Potential places to search include \href{https://github.com/}
{GitHub}, \href{https://swmath.org/} {swMATH} and domain related publications,
such as review articles. Domain Experts are also asked for their suggestions and
are asked to review the initial draft of the software list.

@@

\begin{enumerate}
\item The software functionality must fall within the identified domain.
\item The source code must be viewable.
\item The empirical measures listed in Section \ref{SecEmpiricalMeasures} should
  ideally be available, which implies a preference for GitHub-style
  repositories.
\item The software cannot be marked as incomplete, or in an initial development
  phase.
\end{enumerate}

\section{How to Initially Filter the Software List?} \label{SecInitialFilter}

@@

Copies of both the initial and filtered lists, along with the rationale for
shortening the list, should be kept for traceability purposes.

@@

The final result of the domain analysis will be tables of commonalities,
variabilities, and parameters of variation of a program family.
\cite{smith2008commonality} present a template for conducting a commonality
analysis, which was referred to when conducting this work.
\cite{weiss1998commonality} describes another commonality analysis technique for
deciding the members of a program family. \cite{SmithAndChen2004} and
\cite{SmithMcCutchanAndCarette2017} are examples of a commonality analysis for a
family of mesh generating software and a family of material models,
respectively. The steps to produce a commonality analysis are:

\begin{enumerate}
\item Write an Introduction
\item Write an Overview of the Domain Knowledge
\item List Commonalities
\item List Variabilities
\item List Parameters of Variation
\item Add Terminology, Definitions, Acronyms
\end{enumerate}

% A sample commonality analysis for Lattice Boltzmann Solvers can be found
% \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Peter-Notes/Commonality-Analysis-LB-Systems.pdf}
% {here}.

\section{Repository Based Measures} \label{SecEmpiricalMeasures}

Some quality measurements rely on gathering raw and processed data from software
repositories. We focus on data that is reasonably easy to collect, which we
combine and analyze. The measures that are collected relate to the research
questions (Section~\ref{SecResearchQuestions}). For instance, we collect data to
see how large a project is, to ascertain a project’s popularity, and to
determine whether to project is being actively developed.

Section~\ref{rawdata} lists the raw data that is collected.  Some of this data
can be observed from GitHub repository metrics. The rest can be collected using
freeware tools. \href{https://github.com/tomgi/git_stats}{GitStats} is used to
measure the number of binary files as well as the number of added and deleted
lines in a repository. This tool is also used to measure the number of commits
over different intervals of time. \href{https://github.com/boyter/scc}{Sloc Cloc
and Code (scc)} is used to measure the number of text based files as well as the
number of total, code, comment, and blank lines in a repository. These tools
were selected due to their installability, usability, and ability to gather the
empirical measures listed below. Details on installing and running the tools can
be found in Appendix~\ref{SecRepoTools}.  Section \ref{processeddata} introduces
the required processed data, which is calculated using the raw data.

@@

\begin{itemize}
\item Status of software package as either dead or alive, where alive is defined
  as the presence of repository commits or software package version releases in
  the last 18 months.
\item Percentage of identified issues that are closed.
\item Percentage of code that is comments.
\end{itemize}

\section{Measure Using Measurement Template} \label{SecMeasureTemplate}

@@

The Measurement Template is found in
Appendix~\ref{SecGradingTemplate}.  This template is used to track measurements
and quality scores for all of the software packages in the domain. For each
software package, we fill out one column of the template. This process can take
between 2 to 5 hours for each package.  Project developers can be contacted for
help regarding installation, if necessary, but a cap of about 2 hours should be
imposed on the entire installation process, to keep the overall measurement time
feasible.

@@

\begin{enumerate}
\item Gather the summary information into the top section of the document
  (Figure \ref{measurement_template_image})
\item Using the GitStats tool that is described in Section
  \ref{SecEmpiricalMeasures} gather the measurements for the Repo Metrics -
  GitStats section found near the bottom of the document
\item Using the SCC tool that is also described in Section
  \ref{SecEmpiricalMeasures} gather the measurements for the Repo Metrics - SCC
  section found near the bottom of the document
\item If the software package is found on git, gather the measurements for the
  Repo Metrics - GitHub section found near the bottom of the document
\item Review installation documentation and attempt to install the software
  package on a virtual machine
\item Gather the measurements for installability
\item Gather the measurements for correctness and verifiability
\item Gather the measurements for surface reliability
\item Gather the measurements for surface robustness
\item Gather the measurements for surface usability
\item Gather the measurements for maintainability
\item Gather the measurements for reusability
\item Gather the measurements for surface understandability
\item Gather the measurements for visibility and transparency
\item Assign a score out of ten for each quality. The score can be measured
  using the Measurement Template Impression Calculator, found in
  Appendix~\ref{SecImpressionCalculator}. For each quality measurement, the file
  indicates the appropriate score to assign the measurement based on possible
  measurement values.
\end{enumerate}

@@

The Analytical Hierarchy Process (AHP) is a decision-making technique that can
be used when comparing multiple options by multiple criteria. In our work AHP is
used for comparing and ranking the software packages of a domain using the
quality scores that are gathered in the Measurement Template
(Appendix~\ref{SecGradingTemplate}). AHP performs a pairwise analysis between
each of the quality options using a matrix which is used to generate an overall
score for each software package for the given criteria. \cite{SmithEtAl2016}
shows how AHP is applied to ranking software based on quality measures. We have
developed a
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{tool}
for conducting this process. The tool includes an AHP JAR script and a
sensitivity analysis JAR script that is used to ensure that the software package
rankings are appropriate with respect to the uncertainty of the quality scores.
The
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README
file} outlines the requirements for, and configuration and usage of, the JAR
scripts. The JAR scripts, source code, and required libraries are located in the
same folder as the README file.

% \section{Rank Short List} \label{SecRankShortList}

% An AHP pairwise comparison between the short list packages with respect to
% usability and modifiability experiment results is conducted, and then reviewed
% by the domain expert. The tool can be found here. The usability and
% modifiability survey can be found
% \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}
% {here}. This step is intended to start a conversation about the relative quality
% of short-listed software packages.

\section{Estimate of Time Required} \label{SecEstTimeRequired}

estimate the time required in person hours.  Add a table.  Assume reviewed
material, using provided spreadsheets and AHP tool.  Assume domain has been
identified and domain expert engaged.  Following steps given in
Section~\ref{StepsAQDS}.

\begin{table}[h]
  \caption{Estimated Person Hours for Assessing the State of Practice for Domain
  X} \label{TabPersonHours}
  \centering
  \begin{tabular}{p{10cm} l}
    \toprule
    \textbf{Task} & \textbf{Hours} \\
    \midrule

    Initial 1 hour meeting with domain expert plus meeting prep & 10
    \addtocounter{totHours}{10} \\

    Identify candidate software & 10 \addtocounter{totHours}{10} \\

    Filter software list & 5 \addtocounter{totHours}{5} \\

    Domain analysis & 20 \addtocounter{totHours}{20} \\

    Interaction with domain expert(s) (initial meeting, review software list,
    vet domain analysis, vet AHP ranking) & 10 \addtocounter{totHours}{10} \\

    Collect raw data and fill-in measurement template & 30
    \addtocounter{totHours}{30} \\

    Solicit developers for interviews & 2 \addtocounter{totHours}{2} \\

    Conduct interviews & 15 \addtocounter{totHours}{15} \\

    AHP ranking & 2 \addtocounter{totHours}{2} \\

    Analyze data and answer research questions & 30 \addtocounter{totHours}{30}
    \\

    \midrule
    \textbf{Total} & \textbf{\thetotHours} \\
    \bottomrule
  \end{tabular}

\end{table}

\section{Future Work} \label{SecFutureWork}

- usability and modifiability experiments

Usability experiments (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})

Modifiability experiments (\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Experiments.pdf}{Experiments})

- not just for research software

- use it to inform research on Drasil - what artifacts, what process dev
currently use, how could the new process promoted by Drasil fit into the current
development processes?

- meta analysis of collected data.  Already have GIS, mesh generator.  Currently
working on medical image analysis and LBM.

\section{Concluding Remarks} \label{SecConcludingRemarks}

- summary

@@

\section{A Guide to Repository-Based Measurement Tools} \label{SecRepoTools}
